Topological Eulerian Synthesis of Slow Motion Periodic Videos

### 2018-02-04 Paper/paper.tex
Doing heavy edits on the Laplacian section

% MSB: opportunity to slim paper by cutting examples 
Many prior works use a combination of sliding window embeddings and topological data analysis, such as detecting the Lorenz attractor \cite{de2012topological}, quantifying circadian rhythms in gene expression data \cite{perea2015sliding}, detecting chatter in mechanical systems \cite{khasawneh2016chatter}, detecting wheezing in audio \cite{emrani2014real}, quantifying periodic activites in motion capture data \cite{vejdemo2015cohomological, venkataraman2016persistent}, and quantifying vocal fold anomalies in video \cite{tralie2017quasi}, though we believe we are the first to use a sliding window + TDA framework to {\em synthesize} a new result. Related works have used the graph Laplacian to rearrange images around a loop as a pre-processing step for structure from motion \cite{averbuch2015ringit}, and to reorder the frames of microscope images from a developing embryo around a loop using vector diffusion maps~\cite{dsilva2015diffusionvecordering}.  However, our use of the graph Laplacian to instead {\em parameterize} time series is unique. One can also view the combination of TDA + Laplacian eigenmaps as a simpler alternative to cohomology circular coordinates \cite{de2011persistent,vejdemo2015cohomological} for the special case of periodic data.  It is also the opposite of the approach in \cite{bendich2011improving}, which performs diffusion maps before TDA. \matt{unclear what ''It'' refers to in this last sentence}

@@

The sliding window video embedding \cite{cao1998dynamics,traliehigh,tralie2017quasi} for $\mb{Z}$ can defined as
\begin{equation}
	SW_{D}[\mb{Z}(t)] = \left[ \begin{array}{c} \mb{Z}(t) \\ \mb{Z}(t + 1) \\ \vdots \\ \mb{Z}(t + (D-1))  \end{array} \right] \in \mathbb{R}^{(N-1)\cdot(D)},
\end{equation}
where D is the dimension of the embedding. As shown by Takens~\cite{takens1981detecting}, a sliding window of dimension $2m+1$ of even a single generic observation function of a dynamical system of intrinsic dimension $m$ is sufficient to reconstruct a topological embedding of the underlying trajectory in the original state space.  For periodic signals, this state space is a torus, and $d$ should be twice the number of harmonics present to injectively reconstruct loops on that torus \cite{perea2015sliding}.  The same is true of videos in general \cite{tralie2017quasi}.  Furthermore, the sliding window length $(D-1)$ maximizes the roundness of the embeddings when $(D-1) = k T$ for some integer $k$, where $T$ is the fundamental period.  Since we will use topological data analysis (TDA) to find the scale of the loop, and since TDA measures roundness, it is important that we tune our window size to be an integer multiple of the period.  To do this, we perform 1D ISOMAP \cite{tenenbaum2000global} on the original frames of the video to generate a 1D surrogate signal, and we perform an autocorrelation-based fundamental frequency estimation \cite{Mcleod05asmarter} on this signal to estimate the period and, hence, the appropriate $D$.  This is similar to the diffusion maps + autocorrelation approach in \cite{tralie2017quasi} for analyzing periodic videos.  This estimation often returns integer multiples of the fundamental period, which is fine for our purposes.
% MSB: could probably remove the diffusion maps + autocorrelation approach sentence if we need space

@@

Since sliding window embeddings of periodic videos lie on a topological loop, this means that, at an appropriate scale, a nearest neighbors graph built on the sliding window point cloud will approximately be a graph representing a circle.  To parameterize this graph, we turn to the graph Laplacian \cite{chung1997spectral}. To define the graph Laplacian, we first construct an adjacency matrix $\mb{A}$ in which the entry at $\mb{A}_{ij}$ represents the similarity between windows at times $i$ and $j$: $\mb{A}_{ij} = exp(-||SW_{D}[\mb{Z}(i)] - SW_{D}[\mb{Z}(j)]||_2/\epsilon)$. The graph Laplacian then follows as:
\begin{equation}
	\mb{L} = \mb{D}-\mb{A}
\end{equation}
where $\mb{D}_{ii} = \sum_{j = 1}^N \mb{A}_{ij}, \mb{D}_{i \neq j} = 0$ is the diagonal matrix.  Now let $d_{ij}: V \rightarrow \mathbb{R}^+$ be a nonnegative scalar function on the vertices of $G$.  Then the adjacency matrix correponding to the unweighted graph Laplacain, $A_{\epsilon}^u(d)$, is

\begin{equation}
A_{\epsilon}^u(d)_{ij} = \left\{ \begin{array}{cc} 1 & i \neq j, d_{ij} \leq \epsilon \\ 0 & \text{otherwise} \end{array} \right\}
\end{equation}

Alternatively, the weighted adjacency matrix is defined as $A_{\epsilon}^w(d)_{ij} = \exp(-d_{ij}^2/\epsilon)$ \cite{belkin2003laplacian}, but everything else is the same.  Now suppose we have performed a sliding window $SW_{D}[X[n]]$ for a video $X$, where $X[n] = X[n+kT]$ for some integers $l, k$ so that there are $kT$ points in the sliding window (that is, $SW_{D}[X[i]]$ repeats itself exactly every $T$ frames $k$ times total, for a total of $N = kT$ windows), and that $d_{ij} = ||SW_{D}[X[i]] - SW_{D}[X[j]]||_2$.  In this case, for any $\epsilon > 0$ and $|i - j| = kT$ for some integer $k$, $A^u_{\epsilon}(d)_{ij} = 1$.  Let us further suppose it is possible to choose an $\epsilon$ so that each window is also connected to the window directly before and after it in time, as well as the windows directly before and after its repetitions, but none others; that is, the unweighted adjacency matrix is a symmetric, {\em circulant matrix}, with the first row $A_{\epsilon}^u(d)_{1j} = 1$ if $j = 1, N-1,$ or $lT, 1 \leq l \leq (k-1)$.  Thus, the Laplacian is also circulant with first row

\begin{equation}
\label{eq:modellaplacian}
L_{1j} = \left\{ \begin{array}{cc} 3k-1 & j = 1 \\ -1 & j = lT, 1 \leq l \leq (k-1) \\ -1 & j = 1, N-1 \\ 0 & \text{otherwise} \end{array} \right\}
\end{equation}

@@

The smallest two nonzero eigenvalues occur when $m = k$, corresponding to the eigenvectors \footnote{This generalizes the circle graph used in \cite{averbuch2015ringit}, in which $T = N, k = 1$} 
\begin{equation}
v_{2k}[n] = \cos(2 \pi n / T), v_{2k+1}[n] = \sin(2 \pi n / T)
\end{equation}
When computing the smallest two nonzero eigenvalues numerically, therefore, one coresponds to a sinusoid with period $T$, and the other to an orthogonal sinusoid with period $T$.  When plotted against each other, they form a circle, with an arbitrary phase offset.  Therefore, we compute the circular phase numerically $\phi = \tan^{-1}(\hat{v_1}[n] / \hat{v_2}[n])$, where $\hat{v_i}$ is the eigenvector corresponding to the $i^\text{th}$ smallest nonzero numerical eigenvalue.  In practice, the graphs we get from sliding window may deviate from the ideal Laplacian model in Equation~\ref{eq:modellaplacian}, though they do so gracefully.  Figure~\ref{fig:CirculantExample} shows an example of $\hat{v_1}[n]$ and $\hat{v_2}[n]$ for an ideal graph and for an imperfect graph corresponding to the sliding window embedding of a real video.

Note that for both the unweighted and weighted cases in practice, we have observed that harmonics of the actual frequency of interest occasionally have smaller eigenvalues.  To mitigate this, we use the pair of eigenvectors with the smallest number of zero crossings within the eigenvectors corresponding to the 10 smallest nonzero eigenvalues.

>>>>

% MSB: opportunity to slim paper by cutting examples 
Many prior works use a combination of sliding window embeddings and topological data analysis, such as detecting the Lorenz attractor \cite{de2012topological}, quantifying circadian rhythms in gene expression data \cite{perea2015sliding}, detecting chatter in mechanical systems \cite{khasawneh2016chatter}, detecting wheezing in audio \cite{emrani2014real}, quantifying periodic activites in motion capture data \cite{vejdemo2015cohomological, venkataraman2016persistent}, and quantifying vocal fold anomalies in video \cite{tralie2017quasi}, though we believe we are the first to use a sliding window + TDA framework to {\em synthesize} a new result. Related works have used the graph Laplacian to rearrange images around a loop as a pre-processing step for structure from motion \cite{averbuch2015ringit}, and to reorder the frames of microscope images from a developing embryo around a loop using vector diffusion maps~\cite{dsilva2015diffusionvecordering}.  However, our use of the graph Laplacian to instead {\em parameterize} time series is unique. One can also view the combination of TDA + Laplacian eigenmaps as a simpler alternative to cohomology circular coordinates \cite{de2011persistent,vejdemo2015cohomological} for the special case of periodic data.  Our approach is also the opposite of the approach in \cite{bendich2011improving}, which performs diffusion maps before TDA.

@@

The sliding window video embedding \cite{cao1998dynamics,traliehigh,tralie2017quasi} for $\mb{Z}$ is defined as
\begin{equation}
	SW_{D}[\mb{Z}(t)] = \left[ \begin{array}{c} \mb{Z}(t) \\ \mb{Z}(t + 1) \\ \vdots \\ \mb{Z}(t + (D-1))  \end{array} \right] \in \mathbb{R}^{(N-1)\cdot(D)},
\end{equation}
where D is the dimension of the embedding. As shown by Takens~\cite{takens1981detecting}, a sliding window of dimension $2m+1$ of even a single generic observation function of a dynamical system of intrinsic dimension $m$ is sufficient to reconstruct a topological embedding of the underlying trajectory in the original state space.  For periodic signals, this state space is a torus, and $d$ should be twice the number of harmonics present to injectively reconstruct loops on that torus \cite{perea2015sliding}.  The same is true of videos in general \cite{tralie2017quasi}.  Furthermore, the sliding window length $(D-1)$ maximizes the roundness of the embeddings when $(D-1) = k T$ for some integer $k$, where $T$ is the fundamental period.  Since we will use topological data analysis (TDA) to find the scale of the loop, and since TDA measures roundness, it is important that we tune our window size to be an integer multiple of the period.  To do this, we perform 1D ISOMAP \cite{tenenbaum2000global} on the original frames of the video to generate a 1D surrogate signal, and we perform an autocorrelation-based fundamental frequency estimation \cite{Mcleod05asmarter} on this signal to estimate the period and, hence, the appropriate $D$.  This is similar to the diffusion maps + autocorrelation approach in \cite{tralie2017quasi} for analyzing periodic videos.  This estimation often returns integer multiples of the fundamental period, which is fine for our purposes.
% MSB: could probably remove the diffusion maps + autocorrelation approach sentence if we need space

@@

Since sliding window embeddings of periodic videos lie on a topological loop, this means that, at an appropriate scale, a graph built on the sliding window point cloud will be approximately circular.  We can then compute the sliding window phase $\phi \in [0, 2\pi]$ from this graph using the {\em graph Laplacian} \cite{chung1997spectral}, which we now describe. We first construct an adjacency matrix $\mb{A}$ in which the entry at $\mb{A}_{ij}$ represents the similarity between windows at times $i$ and $j$.  Given a scale $\sigma$, We define an unweighted version as $\mb{A}_{ij}$ is 1 if $||SW_{D}[\mb{Z}(i)] - SW_{D}[\mb{Z}(j)]||_2 \leq \sigma$ and 0 otherwise. We also define a weighted version in which $\mb{A}_{ij} = exp(-||SW_{D}[\mb{Z}(i)] - SW_{D}[\mb{Z}(j)]||_2/2\sigma^2)$. In both cases, we define the graph Laplacian as:
\begin{equation}
	\mb{L} = \mb{D}-\mb{A}
\end{equation}
where $\mb{D}_{ii} = \sum_{j = 1}^N \mb{A}_{ij}, \mb{D}_{i \neq j} = 0$ is the degree matrix representing the sum of all outgoing weights on the diagonal.

We now examine the unweighted Laplacain of an ideal model to motivate our approach.  Suppose we have performed a sliding window $SW_{D}[X[n]]$ for a video $X$ which repeats itself exactly every $T$ frames $k$ times total, for a total of $N = kT$ windows.  Let us further suppose it is possible to choose an $\sigma$ so that the unweighted adjacency matrix contains a 1 entry for the edge between window directly before and after it in time, as well as the windows directly before and after its repetitions, but none others; that is, the Laplacian is a symmetric, {\em circulant} matrix defined as

\begin{equation}
\label{eq:modellaplacian}
L_{ij} = \left\{ \begin{array}{cc} 3k-1 & i = j \\ -1 & |i-j| = lT, l \in \mathbb{Z}^+ \\ -1 & |i-j| = lT \pm 1, l \in \mathbb{Z}^+ \\ 0 & \text{otherwise} \end{array} \right\}
\end{equation}

@@

The smallest two nonzero eigenvalues occur when $m = k$, corresponding to the eigenvectors \footnote{This generalizes the circle graph used in \cite{averbuch2015ringit}, in which $T = N, k = 1$} 
\begin{equation}
v_{2k}[n] = \cos(2 \pi n / T), v_{2k+1}[n] = \sin(2 \pi n / T)
\end{equation}
Therefore, the smallest two numerical nonzero eigenvalues each correspond to sinusoids with period $T$ which are mutually orthogonal.  When plotted against each other, they form a circle with an arbitrary phase offset.  Therefore, we compute the circular phase numerically $\phi = \tan^{-1}(\hat{v_1}[n] / \hat{v_2}[n])$, where $\hat{v_i}$ is the eigenvector corresponding to the $i^\text{th}$ smallest nonzero numerical eigenvalue.  In practice, the graphs we get from sliding window may deviate from the ideal Laplacian model in Equation~\ref{eq:modellaplacian}, though they do so gracefully.  Figure~\ref{fig:CirculantExample} shows an example of $\hat{v_1}[n]$ and $\hat{v_2}[n]$ for the unweighted case for an imperfect graph corresponding to the sliding window embedding of a real video.  To further improve robustness, we default to the weighted Laplacian, since it has nice theoretical roperties for Euclidean point placement matching the adjacency matrix \cite{hall1970r}.  In this case, harmonics of the actual frequency of interest occasionally have smaller eigenvalues when the corresponding harmonics are strongly present in the video.  To mitigate this, we use the pair of eigenvectors with the smallest number of zero crossings within the eigenvectors corresponding to the 10 smallest nonzero eigenvalues.
